Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)

### 2024-03-01 manu-draft.qmd
more revisions

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
```
*Running head*: EBASE metabolic model

@@

Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to unphysical results and poorly constrained errors. Here we present a new approach for estimating estuarine metabolism: EBASE (Estuarine BAyesian Single-station Estimation). EBASE applies a Bayesian framework to a simple process-based model and DO observations, allowing the estimation of critical model parameters, specifically light efficiency and respiration, as informed by a set of prior distributions. EBASE improves upon the stream-based model from which it was derived by accommodating missing DO data and allowing the user to set the time period over which parameters are estimated. We demonstrate that EBASE can recover known metabolic parameters from a synthetic time series, even in the presence of noise (e.g., due to tidal advection) and when prior distributions are uninformed. Optimization periods of 7 days and 30 days are more preferable than 1 day. A comparison with the more-conventional method of Odum reveals the ability of EBASE to avoid unphysical results (such as negative photosynthesis and respiration) and improves when the DO data are detided. EBASE is available using open-source software (R) and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

@@

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, metabolic estimates from EBASE are compared with those of the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$). Both $P$ and $R$ are expressed as positive values, whereas $D$ can be positive (outgassing) or negative (ingassing). The equation has dimensions of moles per unit area and time. $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92]. The gas transfer velocity is modeled using the formulation in @Wanninkhof14:

@@

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$). The optimization period determines how much data are used to estimate each parameter. Only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously given the dependence on other inputs that can vary at each time step (e.g., $PAR$ and wind). The optimization period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to a maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described below. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE. The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria). Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page ([https://fawda123.github.io/EBASE/](#0){.uri}) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a synthesis based on NERRS data [Table 2 in @caffrey2004]. Although this synthesis focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}/\text{m}^2/\text{d}\right)/PAR$. Using the global mean surface $PAR$ of 80 $W/m^2$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}/\text{m}^2$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.

The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004. The standard deviation of $R$ was 171 $\text{mmol}/\text{m}^2/\text{d}$, which was rounded to 50% of the mean, or 150 $\text{mmol}/\text{m}^2/\text{d}$. The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. For the prior standard deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14. However, we chose a larger value, 50%, because the wind speed-based gas transfer velocity formulation of @Wanninkhof14 was mainly developed and calibrated for open-ocean applications, where the complications of surfactants [@frew1997] and tidal currents [@Ho16], as well as fetch and suspended matter [@Borges11], are expected to be minimal.

@@

$$
a \sim N(4, 2)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
$$ {#eq-aprior}

$$
R \sim N(300, 150)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.125)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
$$ {#eq-bprior}

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters. Functions in the EBASE R package allow alternative prior distributions from those above.

@@

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimated by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\text{mmol}/\text{m}^2/\text{d}$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series, 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$. As such, each unique combination of low and high values for the means and standard deviations for $a$ and $R$ were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series, which had a median NSE across all optimization periods and priors of `r prmall[['DO_mod']]`. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons. After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`. This result makes sense, given that differences in $D$ across all optimization periods and priors are due solely to difference in $DO$. Conversely, the $a$ parameter was reproduced the most poorly, with a median NSE of `r prmall[['a']]`.

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck23a] on observed dissolved oxygen time series for Apalachicola Bay in 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series, where the detided results are simply the predictions at average tidal height [see @Beck15 for details]. This method was only applied to dissolved oxygen since it has not been evaluated with any other parameters. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a). Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b. EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck23a]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck23a]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Similar to the previous analyses, the prior distribution for the $b$ parameter was fixed given the difficulty in recovery. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

@@

![The default prior distributions for $a$, $R$, and $b$ used in EBASE. All priors are normal distributions and truncated to positive values. $b$ is also truncated to a maximum of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. Probabilities have been normalized to the maximum probability.](figs/priorplot.png){#fig-priorplot fig-align="center"}

>>>>

```{=tex}
\newcommand{\auni}{\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)}
\newcommand{\runi}{\text{mmol}~\text{m}^{-2}~\text{d}^{-1}}
\newcommand{\buni}{\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)}
```
*Running head*: EBASE metabolic model

@@

Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to estimates that are physically impossible and poorly constrained errors. Here we present a new approach for estimating estuarine metabolism: EBASE (Estuarine BAyesian Single-station Estimation). EBASE applies a Bayesian framework to a simple process-based model and DO observations, allowing the estimation of critical model parameters, specifically light efficiency and respiration, as informed by a set of prior distributions. EBASE improves upon the stream-based model from which it was derived by accommodating missing DO data and allowing the user to set the time period over which parameters are estimated. We demonstrate that EBASE can recover known metabolic parameters from a synthetic time series, even in the presence of noise (e.g., due to tidal advection) and when prior distributions are uninformed. Optimization periods of 7 days and 30 days are more preferable than 1 day. A comparison with the more-conventional method of Odum reveals the ability of EBASE to avoid unphysical results (such as negative photosynthesis and respiration) and improves when the DO data are detided. EBASE is available using open-source software (R) and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries. Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.

@@

The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series. We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series. The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series. Lastly, metabolic estimates from EBASE are compared with those of the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method. These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates. We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study. EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

@@

where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$). Both $P$ and $R$ are expressed as positive values, whereas $D$ can be positive (outgassing) or negative (ingassing). The equation has dimensions of moles per unit area and time (e.g., $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$). $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92]. The gas transfer velocity is modeled using the formulation in @Wanninkhof14:

@@

where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein. Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.

@@

For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>). EBASE differs from BASE in model implementation in two ways. First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$). The optimization period determines how much data are used to estimate each parameter. Only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously given the dependence on other inputs that can vary at each time step (e.g., $PAR$ and wind). The optimization period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to a maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at 1 day. Sensitivity of the EBASE method to different optimization periods is described below. Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters. All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE. The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria). Again, this feature was not included in the original BASE method and a complete time series was required for use. A detailed web page ([https://fawda123.github.io/EBASE/](#0){.uri}) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters. All results provided herein were generated using the EBASE R package (specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>).

@@

Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$ (\~10 $\text{g O}_2~\text{m}^{-2}~\text{d}^{-1}$) using a synthesis based on NERRS data [Table 2 in @caffrey2004]. Although this synthesis focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data. The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/PAR$. Using the global mean surface $PAR$ of 80 $\text{W}~m^{-2}$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}~\text{m}^{-2}$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)$. Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ following @Wanninkhof14.

The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004. The standard deviation of $R$ was 171 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$, which was rounded to 50% of the mean, or 150 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$. The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)$. For the prior standard deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14. However, we chose a larger value, 50%, because the wind speed-based gas transfer velocity formulation of @Wanninkhof14 was mainly developed and calibrated for open-ocean applications, where the complications of surfactants [@frew1997] and tidal currents [@Ho16], as well as fetch and suspended matter [@Borges11], are expected to be minimal.

@@

$$
a \sim N(4, 2)I(0,\infty) \;\; \left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)
$$ {#eq-aprior}

$$
R \sim N(300, 150)I(0,\infty) \;\; \text{mmol}~\text{m}^{-2}~\text{d}^{-1}
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.125)I(0,0.502) \;\; \left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)
$$ {#eq-bprior}

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters. Functions in the EBASE R package allow alternative prior distributions from those above.

@@

The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$. The $b$ parameter was fixed at 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ [@Wanninkhof14]. Daily values of $a$ and $R$ were estimated by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.

@@

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series, 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$. There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$. As such, each unique combination of low and high values for the means and standard deviations for $a$ and $R$ were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series, which had a median NSE across all optimization periods and priors of `r prmall[['DO_mod']]`. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ ([@eq-ebase]). After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`. This result makes sense, given that differences in $D$ across all optimization periods and priors are due solely to difference in $DO$. Conversely, the $a$ parameter was reproduced the most poorly, with a median NSE of `r prmall[['a']]`.

@@

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck23a] on observed dissolved oxygen time series for Apalachicola Bay in 2021. This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series, where the detided results are simply the predictions at average tidal height [see @Beck15 for details]. This method was only applied to dissolved oxygen since it has not been evaluated with any other parameters. Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series. The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a). Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b. EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series. The default priors (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period. The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ following similar logic used for the sensitivity analysis described above.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ [as available in the WtRegDO R package, @Beck23a]. An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results. Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck23a]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE used the default values (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). Similar to the previous analyses, the prior distribution for the $b$ parameter was fixed given the difficulty in recovery. A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.

@@

![The default prior distributions for $a$, $R$, and $b$ used in EBASE. All priors are normal distributions and truncated to positive values. $b$ is also truncated to a maximum of 0.502 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$. Probabilities have been normalized to the maximum probability.](figs/priorplot.png){#fig-priorplot fig-align="center"}
